The role of aneuploidy in the evolution of cancer drug resistance

### 2023-06-19 EvolutionaryRescue1.tex
revised preamble, moved figs and table, fixed bug

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{tikz,pgfplots}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{xurl}

\usepackage{epstopdf}
%\usepackage{listings}
\usepackage{tikz,fp,fp-random,todonotes}
\usepackage[procnames]{listings}
\usepackage{afterpage}
\usepackage{float}
\usepackage{amsthm}
\usepackage{wrapfig}
\usepackage{booktabs,caption}
\usepackage[flushleft]{threeparttable}
\usepackage{scrextend}
\usepackage[all]{nowidow}
\usepackage{stfloats}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{stackengine}
\usepackage[hidelinks]{hyperref}
\usepackage{siunitx}
%\usepackage{subfig}
\usepackage{authblk}
\usepackage{cleveref}

% less space before sections 
% \titlespacing*{<command>}{<left>}{<before-sep>}{<after-sep>}
\usepackage{titlesec}
\titlespacing*{\section}
{0pt}{2ex plus 1ex minus .2ex}{2ex plus .2ex}
\titlespacing*{\subsection}
{0pt}{1ex plus 1ex minus .2ex}{1ex plus .2ex}
\titlespacing*{\paragraph}
{0pt}{1ex plus 1ex minus .2ex}{1ex plus .2ex}

\addtokomafont{labelinglabel}{\sffamily}

\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.text}

\usetikzlibrary{shadings}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.text}
\usetikzlibrary{shapes.multipart}

\allowdisplaybreaks
\renewcommand{\floatpagefraction}{.85}

\newcommand{\x}{{\bf x}}
\renewcommand{\d}{{\rm d}}
\newcommand{\e}{{\rm e}}
\newcommand{\erfc}{{\rm erfc}}
\newcommand{\ii}{{\rm i}}

@@

% appendix
\usepackage[title,page]{appendix}
\usepackage{chngcntr}

@@

% NatBib
%\usepackage[super,comma]{natbib}
%\usepackage[numbers,super,compress]{natbib}
\usepackage[round,colon,authoryear]{natbib}

@@

% this belong in the next paragraph, or remove it altogether
% One mechanism by which aneuploidy has been hypothesized to affect cancer is by providing phenotypic variation and increasing heterogeneity in a tumor.

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}). 
%Thus, the changes in the number of each cell type is described by 
%\begin{equation}
%\begin{aligned}
%w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
%w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
%a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
%a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
%m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
%m_t&\rightarrow m_t-1:\quad \mu_am_t.
%\end{aligned}
%\end{equation}

%%%%%%%%%%%
\begin{figure}[!t]
\centering
\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
\caption{
\textbf{Model illustration.}
\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively. 
Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$. Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$. \textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug. The aneuploid may be tolerant, non-growing, or or partially resistant.
}
\label{figureAneuploidy}
\end{figure}
%%%%%%%%%%%

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Code and data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.

%\newpage
\section*{Results}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Effect of aneuploidy on evolutionary rescue}
To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$),
\begin{equation}\label{ratiorescueexact}
H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Evolutionary rescue time}

@@

\begin{align*}
r_1\left(t\right)&=vp_m\int_0^ta_{\tau}\,\d\tau=\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
r_2\left(t\right)&=vp_m\int_0^tw_{\tau}\,\d\tau=vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}.
\end{align*}
For large initial population sizes we can assume that both rescue mutations produced through direct mutation and aneuploidy are independent and, as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$. Consequently, the mean time to the appearance of the first rescue mutant is: 
\begin{align}\label{meantimet2}
\tau_2=\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)}\,\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{1-\left(1-p_w\right)^N},
\end{align}
which we plot in Figure \ref{MeanTimeGrowthAneuploidyPlot} as a function of the initial population size.

We wish to obtain a simpler formula for $\tau_2$ in an analogous way to \eqref{limitapprox}. For this, we make use of the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),
\end{align*}
which allow us to write:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\approx\frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
\end{equation*}
As a result, the integrand in \eqref{meantimet2} can be written as:
\begin{align*}
&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx\exp\left(-uvNp_m\tau^2-vNp_m\tau\right)\\
&=\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
\end{align*}
Consequently, the mean time $\tau_2$ is obtained to be:
\begin{align}\label{limitapprox2}
\tau_2\approx\left[1+\exp\left(-Np_w\right)\right]\exp\left(\frac{vNp_m}{2u}\right)\frac{\erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
\end{align}
where $\erfc$ is the complementary error function. We plot the expansion \eqref{limitapprox2} in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.

If we select only linear terms in the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2),\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
\end{align*}
we obtain the first order approximation for $\tau_2$:
\begin{align}\label{limitapprox3}
\tau_2\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m},
\end{align}
which we plot in Figure \ref{MeanTimeGrowthAneuploidyPlot} and observe that it offer as a good a fit to \eqref{meantimet2} as \eqref{limitapprox2}. Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors. This is consistent with the information obtained from \Cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Contribution of aneuploidy to mean evolutionary rescue time}
\begin{align}\nonumber
I&=\frac{\tau_2}{\tau_1}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\times\frac{1-\left(1-\left.p_w\right|_{u=0}\right)^N}{1-\left(1-\left.p_w\right|_{u>0}\right)^N}\\\label{eqMeanTimeRatioInitialPopulationSize}
&=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\frac{1}{H},
\end{align}
where $H$, is the ratio of the probability of evolutionary rescue with and without aneuploidy, defined in \eqref{ratiorescueexact}. We plot \eqref{eqMeanTimeRatioInitialPopulationSize} in Figure \ref{MeanTimeRatioInitialPopulationSize} as a function of the initial wildtype population for varying values of the Malthusian fitness of aneuploid cells $\Delta_a$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
{\small
\section*{Acknowledgements}
%We thank XXX for discussions and comments. % TODO who?
This work was supported in part by
the Israel Science Foundation (ISF 552/19, YR),
the US–Israel Binational Science Foundation (BSF 2021276, YR), 
Minerva Stiftung Center for Lab Evolution (YR), 
and the Ela Kodesz Institute for Research on Cancer Development and Prevention (RS).
% TODO add funding for Uri and Daniel
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section*{References}
%\newpage
\nolinenumbers
%\bibliographystyle{unsrtnat}
\bibliographystyle{agsm}
\bibliography{evo2022}
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[p]
% small population size
% A: partial resistance
% B: tolerance
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
\end{subfigure}
\caption{
\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$). % TODO change panel labels to ABCD
When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; (\textbf{Bright}, $N=10^8$), the rescue probability is much lower. 
In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
Here, aneuploidy rate is $u=10^{-2}$ and mutation rate is $v=10^{-7}$.
% TODO add missing parameters
}
\label{rescue_prob_wt_growth}
\end{figure}

@@

\begin{figure}[p]
% small population size
% switch from tolerance to non-growing
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf(A) $N=10^4$; \textbf{(B)} $N=10^8$). 
Markers for simulation results with 95\% CI; 
solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
dashed lines for the approximate formula (\cref{rescue_prob_approx}).
The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
% TODO why no switch in panel B?
% TODO add parameter values
% TODO more simulations for 0 > Delta_a > Delta^*_a
}
\label{rescue_prob_an_growth}
\end{figure}

@@

%\appendix
%\begin{appendices}
%\renewcommand{\theequation}{\thesection\arabic{equation}}
%\counterwithin*{equation}{section}

>>>>

%\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx} % Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
% TeX will automatically convert eps --> pdf in pdflatex
\usepackage[labelfont=bf]{caption}
\usepackage{float}

\usepackage{amssymb,amsmath,amsthm}
\usepackage{commath}
\usepackage[hyphens]{url}
\usepackage[dvipsnames]{xcolor}
\usepackage[unicode=true,colorlinks=true,urlcolor=CadetBlue,citecolor=black,linkcolor=black]{hyperref}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{authblk}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{lipsum}  
\usepackage[title,page]{appendix}
\usepackage{chngcntr}
%\usepackage{end float}
\usepackage{cleveref}
 \usepackage{subcaption}

%SetFonts
% newtxtext+newtxmath
\usepackage{newtxtext} %loads helv for ss, txtt for tt
\usepackage{amsmath}
\usepackage[bigdelims]{newtxmath}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
%SetFonts

% less space before sections 
% https://tex.stackexchange.com/a/101126
%\usepackage{titlesec}
%\titlespacing*{\section}{0pt}{0.5\baselineskip}{0\baselineskip}
%\titlespacing*{\subsection}{0pt}{0.5\baselineskip}{0\baselineskip}

% Species names
%% Meta-Command for defining new species macros
\usepackage{xspace}

\newcommand{\species}[3]{%
  \newcommand{#1}{\gdef#1{\textit{#3}\xspace}\textit{#2}\xspace}}
  \species{\yeast}{Saccharomyces cerevisiae}{S.~cerevisiae}

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\Big[ \stackrel{}{#1}\Big]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
\newtheorem{result}{Result}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{\mathbf{I}}
\newcommand{\E}{\mathbf{E}}
\DeclareMathOperator{\sign}{sign}
\renewcommand{\d}[1]{\ensuremath{\operatorname{d}\!{#1}}}

% Remus commands
\newcommand{\x}{{\bf x}}
\renewcommand{\d}{{\rm d}}
\newcommand{\e}{{\rm e}}
\newcommand{\erfc}{{\rm erfc}}
\newcommand{\ii}{{\rm i}}

@@

@@

% autoref
\def\equationautorefname{Eq.}

% NatBib
\usepackage[comma,sort]{natbib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
We analyze three scenarios: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}; in the third, aneuploid cells are non-growing or "barely growing", that is, either slightly tolerant or slightly resistant, such that $\Delta_a \approx 0$.
Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}).

%Thus, the changes in the number of each cell type is described by 
%\begin{equation}
%\begin{aligned}
%w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
%w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
%a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
%a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
%m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
%m_t&\rightarrow m_t-1:\quad \mu_am_t.
%\end{aligned}
%\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{Code and data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Results}

@@

\subsection*{Effect of aneuploidy on evolutionary rescue}
To determine the extent to which aneuploidy may affect evolutionary rescue, we define $H$ to be the ratio of the rescue probability with aneuploidy ($u>0$) and the rescue probability without aneuploidy ($u=0$),
\begin{equation}\label{ratiorescueexact}
H = \frac{p_{rescue}(u>0)}{p_{rescue}(u=0)}.
\end{equation}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{Evolutionary rescue time}

@@

\begin{align}
r_1\left(t\right)&=vp_m\int_0^ta_{\tau}\,\d\tau=\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\
r_2\left(t\right)&=vp_m\int_0^tw_{\tau}\,\d\tau=vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}.
\end{align}
For large initial population sizes we can assume that both rescue mutations produced through direct mutation and aneuploidy are independent and, as a result, they can be merged into a single Poisson process with rate $\left(r_1+r_2\right)\left(t\right)$. Consequently, the mean time to the appearance of the first rescue mutant is: 
\begin{align}\label{meantimet2}
\tau_2=\frac{\int_0^\infty\e^{-\left(r_1+r_2\right)}\,\d\tau}{1-\left(1-p_w\right)^N}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{1-\left(1-p_w\right)^N},
\end{align}
which we plot in \Cref{MeanTimeGrowthAneuploidyPlot} as a function of the initial population size.

We wish to obtain a simpler formula for $\tau_2$ in an analogous way to \eqref{limitapprox}. For this, we make use of the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+\frac{\Delta_w^2\tau^2}{2}+O(\tau^3)-1}{\Delta_w}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+\frac{\Delta_a^2\tau^2}{2}+O(\tau^3)-1}{\Delta_a}=\tau+\frac{\Delta_w}{2}\tau^2+O(\tau^3),
\end{align*}
which allow us to write:
\begin{equation*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\approx\frac{\left(\Delta_w-\Delta_a\right)\tau^2}{2}.
\end{equation*}
As a result, the integrand in \eqref{meantimet2} can be written as:
\begin{align*}
&\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_w\tau}-1}{\Delta_w}-\frac{\e^{\Delta_a\tau}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\approx\exp\left(-uvNp_m\tau^2-vNp_m\tau\right)\\
&=\exp\left(\frac{vNp_m}{2}\right)\exp\left[-\frac{uvNp_m}{2}\left(\tau+\frac{1}{u}\right)\right].
\end{align*}
Consequently, the mean time $\tau_2$ is obtained to be:
\begin{align}\label{limitapprox2}
\tau_2\approx\left[1+\exp\left(-Np_w\right)\right]\exp\left(\frac{vNp_m}{2u}\right)\frac{\erfc\left(\sqrt{\frac{vNp_m}{2u}}\right)}{\sqrt{\frac{2uvNp_m}{\pi}}},
\end{align}
where $\erfc$ is the complementary error function. We plot the expansion \eqref{limitapprox2} in \Cref{MeanTimeGrowthAneuploidyPlot} and observe that it is a very good fit for large values of the initial wildtype population size.

If we select only linear terms in the following expansions:
\begin{align*}
\frac{\e^{\Delta_w\tau}-1}{\Delta_w}&=\frac{1+\Delta_w\tau+O(\tau^2)-1}{\Delta_w}=\tau+O(\tau^2),\\
\frac{\e^{\Delta_a\tau}-1}{\Delta_a}&=\frac{1+\Delta_a\tau+O(\tau^2)-1}{\Delta_a}=\tau+O(\tau^2),
\end{align*}
we obtain the first order approximation for $\tau_2$:
\begin{align}\label{limitapprox3}
\tau_2\approx\left(1+\e^{-Np_w}\right)\int_0^\infty\e^{-uNp_m\tau}\,\d\tau=\frac{\left(1+\e^{-Np_w}\right)}{uNp_m} ,
\end{align}
which we plot in \Cref{MeanTimeGrowthAneuploidyPlot} and observe that it offer as a good a fit to \cref{meantimet2} as \cref{limitapprox2}. Additionally, we observe that for large initial wildtype populations sizes direct mutation drives evolutionary rescue while aneuploidy plays a role for intermediate sized tumors. This is consistent with the information obtained from \cref{rescue_ratio} where aneuploidy improves the probability of evolutionary rescue only  for small and intermediate values of $N$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Contribution of aneuploidy to mean evolutionary rescue time}
\begin{align}
I&=\frac{\tau_2}{\tau_1}=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-uNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\times\frac{1-\left(1-\left.p_w\right|_{u=0}\right)^N}{1-\left(1-\left.p_w\right|_{u>0}\right)^N}\\\label{eqMeanTimeRatioInitialPopulationSize}
&=\frac{\int_0^\infty\exp\left[-\frac{uvNp_m}{\Delta_w-\Delta_a}\left(\frac{\e^{\Delta_wt}-1}{\Delta_w}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right)-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}\right]\,\d\tau}{\int_0^\infty\e^{-vNp_m\frac{\e^{\Delta_w\tau}-1}{\Delta_w}}\,\d\tau}\frac{1}{H},
\end{align}
where $H$, is the ratio of the probability of evolutionary rescue with and without aneuploidy, defined in \cref{ratiorescueexact}. We plot \cref{eqMeanTimeRatioInitialPopulationSize} in \Cref{MeanTimeRatioInitialPopulationSize} as a function of the initial wildtype population for varying values of the Malthusian fitness of aneuploid cells $\Delta_a$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
{\small
\section*{Acknowledgements}
%We thank XXX for discussions and comments. % TODO who?
This work was supported in part by
the Israel Science Foundation (ISF 552/19, YR),
the US–Israel Binational Science Foundation (BSF 2021276, YR), 
Minerva Stiftung Center for Lab Evolution (YR), 
and the Ela Kodesz Institute for Research on Cancer Development and Prevention (RS).
% TODO add funding for Uri and Daniel
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section*{References}

\nolinenumbers
%\bibliographystyle{unsrtnat}
\bibliographystyle{agsm}
\bibliography{evo2022}

%%%%%%%%%%

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
\caption{
\textbf{Model illustration.}
\textbf{(A)} A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively. 
Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$. Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$. \textbf{(B)} The wildtype and the mutant are susceptible and resistant, respectively, to the drug. The aneuploid may be tolerant, non-growing, or or partially resistant.
}
\label{figureAneuploidy}
\end{figure}

@@

\begin{figure}
% TODO change panel labels to ABCD
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/CombinedSubplotLargePop.pdf}
\end{subfigure}
\caption{
\textbf{Evolutionary rescue probability with partially resistant or tolerant aneuploid cells.}
Rescue probability is very high when aneuploidy provides partial resistance ($\lambda_a=0.01$), in an initially small tumor (\textbf{Aleft}, $N=10^4$) and even more so in an initially large tumor (\textbf{Aright}, $N=10^8$). 
When aneuploidy provides tolerance (\textbf{Bleft}, $N=10^4$; \textbf{Bright}, $N=10^8$), the rescue probability is much lower. 
In both scenarios, rescue probability increase with both the wildtype growth rate (x-axis) and the aneuploidy growth rate (colors).
Markers represent simulation results with 95\% CI; solid and dashed lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob}); dashed lines for the approximate formula (\cref{rescue_prob_approx}), demonstrating that they all agree.
Here, division rate $\lambda_w=\lambda_a=\lambda_m=0.14$, (so that growth rate changes due to variable death rate), mutant death rate $\mu_m=0.13$ (so that mutant growth rate $\Delta_m=0.01$), aneuploidy rate $u=10^{-2}$, mutation rate $v=10^{-7}$.
}
\label{rescue_prob_wt_growth}
\end{figure}
%%%%%%%%%%%

\begin{figure}
% small population size
% switch from tolerance to non-growing
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergence.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=1\textwidth]{Figures/P_est_divergenceLargePopulation.pdf}
\end{subfigure}
\caption{\textbf{Evolutionary rescue probability with tolerant or non-growing aneuploid cells.}
Rescue probability grows with the aneuploid growth rate $\Delta_a$ (x-axis), and is much higher in an initially large tumor than in a small one (\textbf(A) $N=10^4$; \textbf{(B)} $N=10^8$). 
Markers for simulation results with 95\% CI; 
solid lines for the exact formula (\cref{survival_prob} in \cref{rescue_prob});
dashed lines for the approximate formula (\cref{rescue_prob_approx}).
The approximation agrees with the simulation and exact solution when the initial tumor size is large (panel B).
When the tumor size is small (panel A), we switch between the approximation for tolerant and for non-growing aneuploid cells; the switch occurs at $\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}$.
% TODO why no switch in panel B?
% TODO add parameter values
% TODO more simulations for 0 > Delta_a > Delta^*_a
}
\label{rescue_prob_an_growth}
\end{figure}

@@

%\appendix
%\begin{appendices}
%\renewcommand{\theequation}{\thesection\arabic{equation}}
%\counterwithin*{equation}{section}
